---
title: "EDA"
author: "Phyllis Sun"
date: "2025-03-09"
output: html_document
---
```{r}
# Load required libraries
library(ggplot2)
library(tidyr)
library(dplyr)
library(FactoMineR)
library(factoextra)
library(purrr)
library(reshape2)
library(gridExtra)
```


```{r}
#' Read and preprocess the data
#'
#' This script reads processed fish sonar data and prepares it for analysis.
#' The dataset includes fish species, biological characteristics, and frequency response values.
#' @source This function relies on the external script: `read_tidy_export_EVfiles.R`
source('Analysis_Scripts/read_tidy_export_EVfiles.R')

```

# Data Checking ---------------------------------------------------------

```{r}
#' Filter dataset to retain only Lake Trout and Smallmouth Bass
#'
#' @description
#' This dataset contains multiple fish species, but this function filters
#' the data to include only Lake Trout (`lakeTrout`) and Smallmouth Bass (`smallmouthBass`).
#'
#' @return A filtered tibble containing metadata, biological features, and frequency responses.
fish_data <- processed_data %>%
  filter(species %in% c("lakeTrout", "smallmouthBass")) %>%
  select(
    # Fish metadata
    fishNum, species,

    # Biological features
    totalLength, forkLength, weight, girth, dorsoLatHeight, sex,
    airbladderTotalLength, airBladderWidth, airbladderWeight,

    # Frequency response columns
    matches("^F\\d+(\\.\\d+)?$")
  )

```

```{r}
#' Extract frequency-only data
#'
#' @description
#' Creates a dataset containing only the fish number, species, and frequency response columns.
#'
#' @return A tibble with only frequency response data.
freq_only <- fish_data %>%
  select(fishNum, species, matches("^F\\d+(\\.\\d+)?$"))

# Check number of pings per fish
table(fish_data$fishNum)

```
```{r}
#' Identify missing frequency values per fish
#'
#' @description
#' This function calculates the number of missing frequency values for each fish.
#'
#' @return A tibble containing the count of missing values per fish.
missing_per_fish <- freq_only %>%
  group_by(fishNum, species) %>%
  summarise(
    missing_count = sum(is.na(across(starts_with("F")))),
    .groups = "drop"
  ) %>%
  arrange(desc(missing_count))


#' Plot missing frequency values per fish
ggplot(missing_per_fish, aes(x = reorder(fishNum, -missing_count), 
                             y = missing_count, 
                             fill = species)) +
  geom_col() +
  labs(
    title = "Missing Frequency Values Per Fish",
    x = "Fish ID", 
    y = "Number of Missing Values",
    fill = "Species"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  scale_fill_manual(values = c("lakeTrout" = "#1f77b4", "smallmouthBass" = "#ff7f0e"))

```
```{r}
# First identify fish with any missing frequencies
fish_with_missing <- missing_per_fish %>% 
  filter(missing_count > 0) %>% 
  pull(fishNum)

# Calculate frequency of missingness across these fish
missing_freq_dist <- freq_only %>%
  filter(fishNum %in% fish_with_missing) %>%
  summarise(across(starts_with("F"), 
                   ~sum(is.na(.)))) %>%
  pivot_longer(everything(), 
               names_to = "Frequency", 
               values_to = "missing_count") %>%
  mutate(Frequency = as.numeric(gsub("F", "", Frequency))) %>%
  filter(missing_count > 0)  # Remove frequencies with no missing values
# Step 1: Identify missing frequencies for each fish
missing_heatmap_data <- freq_only %>%
  group_by(fishNum, species) %>%
  summarise(across(starts_with("F"), ~any(is.na(.))), .groups = "drop") %>%
  pivot_longer(
    cols = -c(fishNum, species),
    names_to = "Frequency",
    values_to = "is_missing"
  ) %>%
  mutate(Frequency = as.numeric(gsub("F", "", Frequency))) %>%
  filter(is_missing)  # Keep only missing frequencies

# Step 2: Create the heatmap
ggplot(missing_heatmap_data, aes(x = Frequency, y = reorder(fishNum, Frequency))) +
  geom_tile(aes(fill = species), color = "white", linewidth = 0.2) +
  labs(
    title = "Missing Frequencies by Fish",
    x = "Frequency (kHz)",
    y = "Fish ID",
    fill = "Species"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    axis.text.y = element_text(size = 6)
  ) +
  scale_fill_manual(values = c("lakeTrout" = "#1f77b4", "smallmouthBass" = "#ff7f0e")) +
  scale_x_continuous(breaks = seq(45, 260, by = 25))  # Adjust frequency breaks as needed
```
# Data Cleaning ----------------------------------------------------------

```{r}
#' Identify missing values for a specific fish (LT008) in 173–260 kHz range
#'
#' @return A tibble showing the percentage of missing values per frequency.
lt008_missing <- fish_data %>%
  filter(fishNum == "LT008") %>%
  select(matches("^F(17[3-9]|1[89][0-9]|2[0-5][0-9]|260)")) %>%  # Regex for 173–260 kHz
  summarise_all(~sum(is.na(.))) %>%
  pivot_longer(everything(), names_to = "Frequency", values_to = "Missing") %>%
  mutate(Missing_Pct = Missing / nrow(fish_data %>% filter(fishNum == "LT008")) * 100)

#' Remove LT008 if mean missingness >50%
if (mean(lt008_missing$Missing_Pct) > 50) {
  fish_data <- fish_data %>%
    filter(fishNum != "LT008")
}

if (mean(lt008_missing$Missing_Pct) > 50) {
  freq_only <- freq_only %>%
    filter(fishNum != "LT008")
}

```

In the dataset, Lake Trout 008 (LT008) was identified as having more than 50% missing values in the 173–260 kHz frequency range. Based on a threshold of 50% missing data, LT008 was considered an outlier and removed from further analysis to ensure data completeness and reliability.

```{r}
#' Exclude specific frequency ranges
missing_freq <- seq(90, 170, by = 0.5)

freq_clean <- freq_only  %>% 
  select(-any_of(paste0("F", missing_freq)))

fish_clean <- fish_data %>%
  select(-all_of(paste0("F", missing_freq)))
```


# Data Transformation for Visualization -----------------------------------
```{r}
#' Convert frequency and sex values to numeric
freq_clean <- freq_clean %>%
  mutate(across(starts_with("F"), ~ type.convert(.x, as.is = TRUE)))
fish_clean <- fish_clean %>%
  mutate(
    across(starts_with("F"), ~ type.convert(.x, as.is = TRUE)),  # Convert frequency columns
    sex = as.numeric(sex)  # Convert sex to numeric
  )
```

```{r}
#' Transform frequency data into long format for visualization
freq_long <- freq_clean %>%
  select(species, any_of(paste0("F", seq(45, 260, by = 0.5))))%>%
  pivot_longer(
    cols = starts_with("F"),          # Select only frequency columns
    names_to = "Frequency",
    values_to = "Strength",
  ) %>%
  mutate(Frequency_kHz = as.numeric(gsub("F", "", Frequency)))
```


```{r}
#' Compute mean strength per frequency per fish before pivoting
fish_freq_mean_wide <- freq_clean %>%
  group_by(fishNum, species) %>%  
  summarise(across(starts_with("F"), 
                   \(x) mean(x, na.rm = TRUE)), 
            .groups = "drop")

#' Convert to long format for plotting
fish_freq_mean_long <- fish_freq_mean_wide %>%
  select(species, any_of(paste0("F", seq(45, 260, by = 0.5)))) %>%
  pivot_longer(
    cols = starts_with("F"),
    names_to = "Frequency",
    values_to = "Mean_Strength"
  ) %>%
  mutate(Frequency_kHz = as.numeric(gsub("F", "", Frequency)))  # Convert "F45" to 45

```

```{r}
#' Compute mean strength per species per frequency
species_mean <- freq_long %>%
  group_by(species, Frequency_kHz) %>%  # Use numeric Frequency_kHz instead of "Frequency"
  summarize(frequency_mean_strength = mean(Strength, na.rm = TRUE), 
            .groups = "drop")
#' Compute corresponding CI 
species_mean_ci <- freq_long %>%
  group_by(species, Frequency_kHz) %>%
  summarise(
    frequency_mean_strength = mean(Strength, na.rm = TRUE),
    lower = frequency_mean_strength - 1.96 * sd(Strength, na.rm = TRUE) / sqrt(n()),
    upper = frequency_mean_strength + 1.96 * sd(Strength, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

```

```{r, fig.width=20, fig.height=15}

ggplot() +
  # Scatter points for individual fish mean strengths
  geom_point(data = fish_freq_mean_long, 
             aes(x = Frequency_kHz, y = Mean_Strength, color = species), 
             alpha = 0.5, size = 0.5) +  

  # Labels & theme
  labs(title = "Mean Strength per Fish Across Frequencies",
       x = "Frequency (kHz)",
       y = "Mean Strength",
       color = "Fish Species") +
  theme_minimal() +

  # Custom colors for fish species
  scale_color_manual(name = "Fish Species", 
                     values = c("lakeTrout" = "deepskyblue",   
                                "smallmouthBass" = "tomato")) +  

  # Cut the x-axis to hide 90-170 kHz region
  scale_x_continuous(breaks = c(seq(45, 85, by = 10), seq(170, 260, by = 25)),  
                     limits = c(45, 260))  

```

```{r, fig.width=12, fig.height=5}
#' Add a new column to classify frequency ranges
fish_freq_mean_long <- fish_freq_mean_long %>%
  mutate(Frequency_Range = case_when(
    Frequency_kHz >= 45 & Frequency_kHz <= 90 ~ "45-90 kHz",
    Frequency_kHz >= 173 & Frequency_kHz <= 260 ~ "173-260 kHz",
    TRUE ~ NA_character_  # Exclude middle range
  )) %>%
  filter(!is.na(Frequency_Range))  # Remove NA values (90-170 kHz)

fish_mean_long <- fish_freq_mean_long %>%
  mutate(Frequency_Range = factor(Frequency_Range, 
                                  levels = c("45-90 kHz", "173-260 kHz")))

#' Plot two frequency range separately 
ggplot(fish_freq_mean_long, aes(x = Frequency_kHz, 
                                y = Mean_Strength, 
                                color = species)) +
  geom_point(size = 0.8,alpha = 0.7) +  # Individual fish trends
  facet_wrap(~ Frequency_Range, scales = "free_x") +  # Separate panels for each frequency range
  labs(
    title = "Mean Strength per Fish Across Two Frequency Ranges",
    x = "Frequency (kHz)",
    y = "Mean Strength",
    color = "Fish Species"
  ) +
  theme_minimal()
```

```{r}
#' Compute mean and 95% confidence interval
#'
#' @param x Numeric vector.
#' @param confidence Confidence level (default: 0.95).
#' @return A tibble with mean, lower, and upper confidence bounds.
compute_ci <- function(x, confidence = 0.95) {
  n <- sum(!is.na(x))  # Ensure NA values don't count in length
  mean_x <- mean(x, na.rm = TRUE)
  stderr <- sd(x, na.rm = TRUE) / sqrt(n)  # Standard Error
  
  # Avoid calculation errors if n < 2
  if (n < 2) {
    return(tibble(mean = NA, lower = NA, upper = NA))
  }
  
  error_margin <- qt(1 - (1 - confidence) / 2, df = n - 1) * stderr  # t-score * SE
  
  tibble(
    mean = mean_x,
    lower = mean_x - error_margin,
    upper = mean_x + error_margin
  )
}

# Get unique fish IDs
labelled_fish <- unique(freq_clean$fishNum)

#' Generate and print confidence interval plots for each fish
plot_fish_ci <- function(fish_id) {
  one_fish_data <- freq_clean %>%
    filter(fishNum == fish_id) %>%
    select(species, any_of(paste0("F", seq(45, 260, by = 0.5)))) %>%
    pivot_longer(
      cols = starts_with("F"),
      names_to = "Frequency",
      values_to = "Strength"
    ) %>%
    mutate(Frequency = as.numeric(gsub("F", "", Frequency)))  # Convert "F45" to 45
  
  # Compute mean and confidence intervals per frequency
  ci_data <- one_fish_data %>%
    group_by(Frequency) %>%
    summarise(compute_ci(Strength), .groups = "drop") %>%
    arrange(Frequency)
  
  # Generate plot
  ggplot(ci_data, aes(x = Frequency, y = mean)) +
    geom_line(color = "blue") +  
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3, fill = "blue") +  
    labs(
      title = paste(fish_id, "Mean Strength with 95% Confidence Interval"),
      x = "Frequency (Hz)",
      y = "Strength"
    ) +
    theme_minimal()
}

#' Generate plots for all fish in parallel using `walk()`
walk(labelled_fish, ~ print(plot_fish_ci(.x)))
```

```{r}
#' Compute total number of unique frequencies per fish
n_frequencies <- length(c(seq(45, 89.5, by = 0.5), seq(173, 260, by = 0.5)))

#' Count number of observations per fish
fish_counts <- freq_clean %>%
  count(fishNum)

#' Repeat fishNum based on pre-pivot row counts
fishNum_vector <- rep(fish_counts$fishNum, 
                      times = fish_counts$n * n_frequencies)  # Repeat for each fish

#' Ensure correct length before assigning fishNum back
if (length(fishNum_vector) == nrow(freq_long)) {
  freq_long_with_id <- freq_long
  freq_long_with_id$fishNum <- fishNum_vector
} else {
  stop("Mismatch in row counts: Check n_frequencies calculation and fish observations")
}

#' Add fishNum back
freq_long_with_id <- freq_long 
freq_long_with_id$fishNum <- fishNum_vector
```

```{r}
#' Detect outliers per frequency per fish
outliers_per_fish <- freq_long_with_id %>% 
  group_by(fishNum, Frequency) %>%  # Compute outliers PER frequency PER fish
  mutate(
    Q1 = quantile(Strength, 0.25, na.rm = TRUE),
    Q3 = quantile(Strength, 0.75, na.rm = TRUE),
    IQR = Q3 - Q1,
    Lower_Bound = Q1 - 1.5 * IQR,
    Upper_Bound = Q3 + 1.5 * IQR,
    Outlier = Strength < Lower_Bound | Strength > Upper_Bound
  ) %>%
  ungroup()

#' Extract only outliers
outliers_detected <- outliers_per_fish %>%
  filter(Outlier == TRUE)

#' Print outlier details
print(outliers_detected)
```

```{r}
#' Define a threshold for rare outliers
outlier_percentage <- 0.0005  

#' Count total observations per fish
fish_observation_counts <- freq_long_with_id %>%
  count(fishNum, name = "total_obs")

#' Count outliers per fish per frequency
outliers_summary <- outliers_per_fish %>%
  filter(Outlier) %>%
  count(fishNum, Frequency, name = "outlier_count") %>%
  left_join(fish_observation_counts, by = "fishNum") %>%
  mutate(threshold = total_obs * outlier_percentage)  # Compute threshold dynamically

print(outliers_summary)

#' Identify outliers that are "rare" based on the threshold
low_outlier_frequencies <- outliers_summary %>%
  filter(outlier_count <= threshold) %>%
  select(fishNum, Frequency)

#' Remove only rare outliers
clean_freq_data <- outliers_per_fish %>%
  anti_join(
    low_outlier_frequencies %>% mutate(Outlier = TRUE), 
    by = c("fishNum", "Frequency", "Outlier")
  )
print(clean_freq_data)


```

```{r}
#' Compute mean strength per frequency per fish BEFORE outlier removal
mean_strength_before <- freq_long_with_id %>%
  group_by(fishNum, Frequency) %>%
  summarise(
    mean_strength = mean(Strength, na.rm = TRUE), 
    .groups = "drop"
  )

#' Compute mean strength per frequency per fish AFTER outlier removal
mean_strength_after <- clean_freq_data %>%
  group_by(fishNum, Frequency) %>%
  summarise(
    mean_strength = mean(Strength, na.rm = TRUE), 
    .groups = "drop"
  )

#' Compute the difference in mean strength before and after outlier removal
mean_difference <- mean_strength_before %>%
  inner_join(mean_strength_after, 
             by = c("fishNum", "Frequency"), 
             suffix = c("_before", "_after")) %>%
  mutate(change = mean_strength_after - mean_strength_before) %>%
  arrange(desc(abs(change)))  # Sort by largest absolute change

#' Print summary of mean differences
print(mean_difference)
```

```{r}

#' Plot the distribution of changes in mean strength after outlier removal
#'
#' This histogram visualizes the distribution of differences in mean strength 
#' per frequency per fish before and after outlier removal.
ggplot(mean_difference, aes(x = change)) +
  geom_histogram(
    bins = 50, 
    fill = "blue", 
    alpha = 0.5
  ) +
  labs(
    title = "Distribution of Changes in Mean Strength Per Frequency Per Fish",
    x = "Change in Mean Strength (After - Before)",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Centered bold title
    axis.title.x = element_text(face = "bold"),  # Bold x-axis label
    axis.title.y = element_text(face = "bold")   # Bold y-axis label
  )

```
Actually we can see no big change, most changes are near 0. So, I decided to keep these outliers. 
```{r}

# Select only frequency response columns for PCA
pca_data <- freq_clean %>%
  select(any_of(paste0("F", seq(45, 260, by = 0.5)))) %>%
  drop_na()  # Remove missing values

# Perform PCA
pca_result <- PCA(pca_data, scale.unit = TRUE, graph = FALSE)

# Scree plot: Variance explained by each principal component
fviz_eig(pca_result, addlabels = TRUE, ylim = c(0, 50))

# Extract PC scores for plotting
pca_scores <- as.data.frame(pca_result$ind$coord) %>%
  mutate(species = freq_clean$species)  # Add species labels back

# Extract explained variance from PCA results
explained_variance <- pca_result$eig[, 2]  # Get percentage of variance explained
pc1_var <- round(explained_variance[1], 2)  # Variance explained by PC1
pc2_var <- round(explained_variance[2], 2)  # Variance explained by PC2

# PCA scatter plot with percentage on axes
ggplot(pca_scores, aes(x = Dim.1, y = Dim.2, color = species)) +
  geom_point(alpha = 0.7) +
  labs(title = "PCA of Frequency Response Data",
       x = paste0("Principal Component 1 (PC1) - ", pc1_var, "% Variance"),
       y = paste0("Principal Component 2 (PC2) - ", pc2_var, "% Variance"),
       color = "Species") +
  theme_minimal()
```

PC1 explains 48.59% of the variance, so it contains meaningful information, but it's not fully separating species.
PC2 explains only 6.93%, which is quite low—this suggests PC2 might not be informative for classification.
The overlapping points indicate that frequency response patterns between species are not easily distinguished using just these two components.

```{r}
# Extract PCA loadings
loadings <- as.data.frame(pca_result$var$coord)  # Get the PCA variable coordinates

# Sort and display top contributing features for PC1 and PC2
top_PC1 <- loadings %>%
  arrange(desc(abs(Dim.1))) %>%
  head(10)  # Top 10 most important frequencies for PC1

top_PC2 <- loadings %>%
  arrange(desc(abs(Dim.2))) %>%
  head(10)  # Top 10 most important frequencies for PC2

# Print the most influential frequency variables
print("Top contributing frequencies for PC1:")
print(top_PC1)

print("Top contributing frequencies for PC2:")
print(top_PC2)

```
PCA does not work well due to the overlapping. 

```{r}
#' Plot PCA Biplot: Frequency Contributions & Species Clusters
#'
#' - Displays frequency contributions as variable arrows.
#' - Colors variables based on contribution.
#' - Makes individual fish data less prominent.
#' - Uses `repel = TRUE` to avoid text overlap.
fviz_pca_biplot(
  pca_result,
  label = "var",       # Show frequency labels
  col.var = "contrib", # Color by contribution
  col.ind = "gray",    # Make fish data less prominent
  repel = TRUE         # Avoid overlapping text
) +
  labs(
    title = "PCA Biplot: Frequency Contributions & Species Clusters",
    x = "Principal Component 1 (PC1)",  # Explicit axis labels
    y = "Principal Component 2 (PC2)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),  # Centered & bold title
    axis.title.x = element_text(face = "bold"),  # Bold x-axis label
    axis.title.y = element_text(face = "bold")   # Bold y-axis label
  )
```
From this visualization, we observe that the most important frequencies fall within the 63–77 kHz range, contributing significantly to PC1, which explains 48.6% of the total variance. These frequencies likely capture broad variations in fish frequency responses, but their role in distinguishing species remains uncertain.

```{r}
# Define a function to compute and plot correlation heatmaps per species
plot_species_corr <- function(species_name, data) {
  # Filter data for the selected species
  species_data <- data %>% filter(species == species_name) %>%
    select(any_of(paste0("F", seq(45, 260, by = 0.5))))
  
  # Compute correlation matrix
  corr_matrix <- cor(species_data, use = "pairwise.complete.obs", method = "pearson")
  
  # Convert to long format for plotting
  corr_long <- melt(corr_matrix)
  
  # Generate heatmap
  ggplot(corr_long, aes(x = Var1, y = Var2, fill = value)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    labs(title = paste("Heatmap of Frequency Correlations for", species_name),
         x = "Frequency (kHz)",
         y = "Frequency (kHz)",
         fill = "Correlation") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
      plot.title = element_text(size = 8, face = "bold")  # Adjust title size here
    )
}
# Generate heatmaps separately for Lake Trout and Smallmouth Bass
p1 <- plot_species_corr("lakeTrout", freq_clean)
p2 <- plot_species_corr("smallmouthBass", freq_clean)
grid.arrange(p1, p2, ncol = 2)
```
Since the frequency is obviously not independent of each other I decide not to use Bonferroni correction. 

```{r}
# Function to compute highly correlated frequencies for a given species
identify_highly_correlated_frequencies <- function(data, species_name, threshold = 0.7) {
  # Filter data for the given species
  species_data <- data %>%
    filter(species == species_name) %>%
    select(any_of(paste0("F", seq(45, 260, by = 0.5))))

  # Compute correlation matrix
  corr_matrix <- cor(species_data, use = "pairwise.complete.obs")

  # Convert correlation matrix into a long format
  corr_long <- as.data.frame(as.table(corr_matrix)) %>%
    rename(Frequency1 = Var1, Frequency2 = Var2, Correlation = Freq) %>%
    filter(Frequency1 != Frequency2 & abs(Correlation) > threshold)  # Filter for strong correlations

  # Remove duplicate pairs (e.g., F45-F50 is the same as F50-F45)
  corr_long <- corr_long %>%
    rowwise() %>%
    mutate(FreqPair = paste(sort(c(Frequency1, Frequency2)), collapse = "-")) %>%
    distinct(FreqPair, .keep_all = TRUE) %>%
    select(-FreqPair)

  return(corr_long)
}

# Identify highly correlated frequencies for each species
corr_tbl_lt <- highly_corr_LakeTrout <- identify_highly_correlated_frequencies(freq_clean, "lakeTrout", threshold = 0.7)
corr_tbl_smb <- highly_corr_SmallmouthBass <- identify_highly_correlated_frequencies(freq_clean, "smallmouthBass", threshold = 0.7)

# Display top correlated frequency pairs for each species
print("Highly Correlated Frequencies (Lake Trout):")
print(head(corr_tbl_lt, 10))

print("Highly Correlated Frequencies (Smallmouth Bass):")
print(head(corr_tbl_smb, 10))

```

```{r}
# Function to compute correlations of feature for each species
compute_feature_correlation <- function(species_name) {
  species_data <- fish_clean %>%
    filter(species == species_name) %>%  # Filter species
    mutate(sex = as.numeric(sex)) %>%
    select(sex, totalLength, forkLength, weight, girth, 
           airbladderTotalLength, airBladderWidth, airbladderWeight, 
           matches("^F\\d+(\\.\\d+)?$"))  # Select bio features + frequencies
  # Compute correlation matrix
  cor_matrix <- cor(species_data, use = "pairwise.complete.obs")

  # Convert to long format for visualization
  cor_long <- as.data.frame(as.table(cor_matrix)) %>%
    rename(Variable1 = Var1, Variable2 = Var2, Correlation = Freq) %>%
    filter(Variable1 %in% c("sex", "totalLength", "forkLength", "weight", "girth", 
                            "airbladderTotalLength", "airBladderWidth", "airbladderWeight") &
           grepl("^F", Variable2))  # Keep only correlations between bio traits & frequencies
  
  cor_long$species <- species_name  # Add species label
  return(cor_long)
}

# Compute correlations separately
cor_lake_trout <- compute_feature_correlation("lakeTrout")
cor_smallmouth_bass <- compute_feature_correlation("smallmouthBass")

# Combine results
species_corr <- bind_rows(cor_lake_trout, cor_smallmouth_bass)
species_corr
```

```{r}
# Set correlation threshold for counting high correlations
cor_threshold <- 0.4  

# Process each species separately
species_summary <- species_corr %>%
  group_by(species, Variable1) %>%
  summarise(
    Max_Correlation = max(abs(Correlation), na.rm = TRUE),  # Max absolute correlation
    Mean_Correlation = mean(abs(Correlation), na.rm = TRUE),  # Mean absolute correlation
    Frequency_Count = sum(abs(Correlation) > cor_threshold, na.rm = TRUE),  # Count of high correlations
    Max_Correlation_Frequency = Variable2[which.max(abs(Correlation))],  # Frequency with max correlation
    .groups = "drop"
  ) %>%
  arrange(species, desc(Max_Correlation))  # Sort by species & descending max correlation

# View the cleaned summary table
species_summary

```

```{r}
#' Prepare correlation data by extracting frequency and binning
species_corr <- species_corr %>%
  mutate(
    Frequency_kHz = as.numeric(gsub("^F", "", Variable2)),  # Extract numeric frequency
    Frequency_Bin = round(Frequency_kHz / 10) * 10          # Bin frequencies by 10 kHz
  ) %>%
  group_by(species, Variable1, Frequency_Bin) %>%
  summarise(
    Mean_Correlation = mean(Correlation, na.rm = TRUE), 
    .groups = "drop"
  )

#' Plot heatmap of species-specific correlations between biological features and frequency response
ggplot(species_corr, aes(x = Frequency_Bin, y = Variable1, fill = Mean_Correlation)) +
  geom_tile() +
  facet_wrap(~species) +  # Separate heatmap per species
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red", midpoint = 0
  ) +
  labs(
    title = "Species-Specific Correlation Between Biological Features and Frequency Response",
    x = "Frequency (kHz, Binned by 10)",
    y = "Biological Features",
    fill = "Correlation"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10, face = "bold", hjust = 0.5),  # Adjust title size, bold, and center
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 10, face = "bold")  # Adjust facet labels
  )

```
The heatmap reveals species-specific patterns in the correlation between biological features and frequency response. For Lake Trout, air bladder width exhibits the strongest correlation, particularly around F67, suggesting that this feature plays a dominant role in shaping the acoustic response of the species. Other features such as weight and girth show moderate but consistent correlations across multiple frequencies, indicating a broader but less pronounced influence. In contrast, for Smallmouth Bass, the strongest correlation is observed with girth, weight, and fork length, particularly around F174, suggesting that body shape and size have a greater impact on the species' frequency response patterns than air bladder width. These differences highlight distinct acoustic signatures driven by species morphology, which are critical for classification tasks. The heatmap is particularly useful for detecting broad trends, ensuring that key relationships are not missed when selecting frequency bands for predictive modeling.

```{r}
# Select relevant biological features
bio_features <- c("totalLength", "forkLength", "weight", "girth", "sex",
                  "airbladderTotalLength", "airBladderWidth", "airbladderWeight")

# Convert data to long format for easier visualization
bio_long <- fish_clean %>%
  select(species, all_of(bio_features)) %>%
  pivot_longer(-species, names_to = "Feature", values_to = "Value")

# Plot boxplots by species
ggplot(bio_long, aes(x = species, y = Value, fill = species)) +
  geom_boxplot(alpha = 0.6, outlier.colour = "red", outlier.size = 1) +
  facet_wrap(~Feature, scales = "free_y") +
  labs(title = "Boxplots of Biological Features by Species",
       x = "Species", y = "Value") +
  theme_minimal() +
  theme(legend.position = "none", strip.text = element_text(size = 5, face = "bold"))

```

```{r}
# Define biological features of interest
bio_features <- c("totalLength", "forkLength", "weight", "girth", 
                  "airbladderTotalLength", "airBladderWidth", "airbladderWeight")

# Filter out invalid sex values (if any) and pivot to long format
df_sex_long <- fish_clean %>%
  filter(sex %in% c("2", "1")) %>%    
  select(species, sex, all_of(bio_features)) %>%
  pivot_longer(cols = -c(species, sex), names_to = "Feature", values_to = "Value")

# Boxplots by sex within each species
ggplot(df_sex_long, aes(x = sex, y = Value, fill = sex)) +
  geom_boxplot(alpha = 0.7, outlier.colour = "red") +
  facet_grid(species ~ Feature, scales = "free_y") +  # separate species by rows, features by columns
  labs(
    title = "Comparison of Biological Features by Sex Within Each Species",
    x = "Sex",
    y = "Value"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
# correlation threshold
cor_threshold <- 0.5  

# Biological + frequency columns
bio_freq_cols <- c("totalLength", "forkLength", "weight", "girth", 
                   "airbladderTotalLength", "airBladderWidth", "airbladderWeight",
                   grep("^F\\d+(\\.\\d+)?$", names(fish_clean), value = TRUE))

# Function to compute correlation for a given species & sex
compute_sex_corr <- function(species_name, sex_label) {
  # Filter data
  subset_data <- fish_clean %>%
    filter(species == species_name, sex == sex_label) %>%
    select(all_of(bio_freq_cols)) %>%
    mutate(across(everything(), as.numeric))  # Ensure numeric

  # Compute correlation matrix
  cor_matrix <- cor(subset_data, use = "pairwise.complete.obs")

  # Convert to long format
  cor_df <- as.data.frame(as.table(cor_matrix)) %>%
    rename(Variable1 = Var1, Variable2 = Var2, Correlation = Freq) %>%
    filter(Variable1 %in% bio_features & grepl("^F", Variable2))  # Only keep bio vs freq

  cor_df$species <- species_name
  cor_df$sex <- sex_label
  return(cor_df)
}

corr_lt_male <- compute_sex_corr("lakeTrout", "2")
corr_lt_female <- compute_sex_corr("lakeTrout", "1")
all_sex_corr_lt <- rbind(corr_lt_male, corr_lt_female)
print(all_sex_corr_lt %>% arrange(desc(Correlation))) 

corr_smb_male <- compute_sex_corr("smallmouthBass", "2")
corr_smb_female <- compute_sex_corr("smallmouthBass", "1")
all_sex_corr_smb <- rbind(corr_smb_male, corr_smb_female)
print(all_sex_corr_smb %>% arrange(desc(Correlation)))
```
This is not focus on the project objective so I decide not to use this plot. 
