---
title: "Modelling"
author: "Phyllis Sun"
date: "2025-03-09"
output: html_document
---
```{r,warning=FALSE}
library(SuperLearner)  # Core SuperLearner package
library(caret)         # Model training and evaluation
library(glmnet)        # Lasso/ElasticNet Regression (SL.glmnet)
library(randomForest)  # Random Forest (SL.randomForest)
library(xgboost)       # Gradient Boosted Trees (SL.xgboost)
library(ranger)        # Fast Random Forest (SL.ranger)
library(nnet)          # Neural Networks (SL.nnet)
library(pROC)          # ROC analysis
library(cvAUC)         # Cross-validated AUC calculation
```

# Building Function for Splitting Data into Train and Test ---------------------

```{r}
#' Split dataset into training and test sets
#'
#' @description
#' This function partitions the dataset into 80% training and 20% test data 
#' while ensuring class balance for each species.
#'
#' @param data A tibble containing fish data.
#' @return A list containing `train_data` and `test_data`.
split_data <- function(data) {
  set.seed(555)
  train_list <- list()
  test_list <- list()
  
  for (sp in unique(data$species)) {
    species_data <- filter(data, species == sp)
    unique_fish <- unique(species_data$fishNum)
    
    train_fish <- sample(unique_fish, size = round(length(unique_fish) * 0.8))
    train_data_sp <- filter(species_data, fishNum %in% train_fish)
    test_data_sp  <- filter(species_data, !fishNum %in% train_fish)
    
    train_list[[sp]] <- train_data_sp
    test_list[[sp]]  <- test_data_sp
  }
  
  list(
    train_data = bind_rows(train_list) %>% sample_frac(1),
    test_data  = bind_rows(test_list) %>% sample_frac(1)
  )
}

```


```{r}
#' Prepare input (X) and output (Y) datasets
#'
#' @description
#' This function extracts features (X) and binary target labels (Y).
#'
#' @param train_data The training dataset.
#' @param test_data The test dataset.
#' @return A list with `X_train`, `X_test`, `Y_train`, `Y_test`.
prepare_xy <- function(train_data, test_data, feature_cols = NULL) {
  if (is.null(feature_cols)) {
    # By default, select all features except species and fishNum
    X_train <- select(train_data, -species, -fishNum)
    X_test  <- select(test_data, -species, -fishNum)
  } else {
    # If feature_cols provided, select those columns
    X_train <- select(train_data, all_of(feature_cols)) %>% select(-species, -fishNum)
    X_test  <- select(test_data, all_of(feature_cols)) %>% select(-species, -fishNum)
  }
  
  Y_train <- as.numeric(factor(train_data$species, levels = c("lakeTrout", "smallmouthBass"))) - 1
  Y_test  <- as.numeric(factor(test_data$species, levels = c("lakeTrout", "smallmouthBass"))) - 1
  list(X_train = X_train, X_test = X_test, Y_train = Y_train, Y_test = Y_test)
}


```

# Buiding Function for Training SuperLearner Model -----------------------------

```{r}
#' Train SuperLearner Model with Optional Meta-Learner Method
#'
#' @description
#' This function trains a SuperLearner ensemble model using multiple base learners:
#' GLMnet (elastic net logistic regression), Random Forest, XGBoost, Neural Network,
#' and Ranger (fast random forest). The ensemble weights are optimized using a specified
#' meta-learning method (default: method.AUC for AUC optimization).
#'
#' @param X_train A data frame or tibble containing the training feature matrix.
#' @param Y_train A numeric or factor vector containing binary target labels for training.
#'   The labels should be coded as 0 (lakeTrout) and 1 (smallmouthBass).
#' @param meta_method A character string specifying the meta-learning method to use.
#'   Defaults to "method.AUC". Other options include "method.NNLS" or "method.NNloglik", etc.
#'
#' @return A trained SuperLearner model object.
train_superlearner <- function(X_train, Y_train, meta_method = "method.AUC") {
  learners <- c("SL.glmnet", "SL.randomForest", "SL.xgboost", "SL.nnet", "SL.ranger")
  
  SuperLearner(
    Y = as.numeric(Y_train), 
    X = X_train, 
    family = binomial(),
    SL.library = learners,
    method = meta_method,  
    cvControl = list(V = 5)   # 5-fold cross-validation
  )
}

```

## Model Training on Six Dataset Variants 

# 1. All Biological Features + All Frequencies----------------------------------

```{r}
# Execute data split
split_results1 <- split_data(fish_clean)
train_data <- split_results1$train_data
test_data  <- split_results1$test_data

# check species balance
cat("Training set species distribution:\n")
print(table(train_data$species))
cat("Test set species distribution:\n")
print(table(test_data$species))

# Prepare input (X) and output (Y) datasets
xy_data <- prepare_xy(train_data, test_data)
X1_train <- xy_data$X_train
X1_test  <- xy_data$X_test
Y1_train <- xy_data$Y_train
Y1_test  <- xy_data$Y_test
```


```{r}
# Train the SuperLearner model on All bio + All freq data
SL1 <- train_superlearner(X1_train, Y1_train)
summary(SL1)

# Extract learner weights and CV risk into a summary dataframe
sl1_summary <- data.frame(
  Algorithm = names(SL1$coef),   # Extract learner names
  Weight = SL1$coef,             # Learner weights
  CV_Risk = SL1$cvRisk           # Cross-validation risk per learner
)

# View the result
print(sl1_summary)
```
In cases where SuperLearner assigns uniform weights across all base learners, we default to XGBoost for its proven robustness and prior strong performance on similar data variants.

```{r}
#' Train and evaluate XGBoost model
#'
#' @description
#' This function trains an XGBoost model and computes balanced accuracy, confusion matrix, AUC-ROC, and F1 score.
#'
#' @param X_train Training feature matrix.
#' @param Y_train Training labels.
#' @param X_test Test feature matrix.
#' @param Y_test Test labels.
#' @return A list including confusion matrix, balanced accuracy, AUC, and F1 score.

train_xgboost <- function(X_train, Y_train, X_test, Y_test) {
  X_train_mat <- as.matrix(X_train)
  X_test_mat  <- as.matrix(X_test)
  
  xgb_model <- xgboost(data = X_train_mat, label = Y_train, 
                       objective = "binary:logistic", nrounds = 200, verbose = 0)
  
  predictions <- predict(xgb_model, newdata = X_test_mat)
  final_predictions <- ifelse(predictions > 0.5, 1, 0)
  
  accuracy <- mean(final_predictions == Y_test) * 100
  message("XGBoost Test Accuracy: ", round(accuracy, 2), "%")
  
  # Confusion matrix
  conf_matrix <- confusionMatrix(as.factor(final_predictions), as.factor(Y_test))
  print(conf_matrix)
  
  # Balanced Accuracy
  sensitivity_value <- conf_matrix$byClass["Sensitivity"]
  specificity_value <- conf_matrix$byClass["Specificity"]
  balanced_accuracy <- (sensitivity_value + specificity_value) / 2
  message("Balanced Accuracy: ", round(balanced_accuracy, 4))
  
  # AUC-ROC score
  roc_curve <- roc(Y_test, predictions)
  auc_value <- auc(roc_curve)
  message("AUC-ROC Score: ", round(auc_value, 4))
  
  # F1 score calculation
  precision <- conf_matrix$byClass["Pos Pred Value"]
  recall <- sensitivity_value
  f1_score <- 2 * (precision * recall) / (precision + recall)
  message("F1 Score: ", round(f1_score, 4))
  
  return(list(
    ConfusionMatrix = conf_matrix,
    BalancedAccuracy = balanced_accuracy,
    AUC = auc_value,
    F1_Score = f1_score
  ))
}
```

```{r}
# Show the evaluation on XGBoost result for All frequency + All bio data
data1_xg_result <- train_xgboost(X1_train, Y1_train, X1_test, Y1_test)
```

```{r}
#' Train and evaluate Random Forest model
#'
#' @description
#' This function trains a Random Forest model and evaluates its performance using 
#' confusion matrix, balanced accuracy, AUC-ROC, and F1 score.
#'
#' @param X_train Training feature matrix.
#' @param Y_train Training labels.
#' @param X_test Test feature matrix.
#' @param Y_test Test labels.
#' @return A list including confusion matrix, balanced accuracy, AUC, and F1 score.

train_random_forest <- function(X_train, Y_train, X_test, Y_test) {
  
  # Train model (Y must be factor for classification)
  Y_train_factor <- as.factor(Y_train)
  rf_model <- randomForest(x = X_train, y = Y_train_factor, ntree = 200, importance = TRUE)
  
  # Predictions
  rf_predictions <- predict(rf_model, newdata = X_test)
  rf_probabilities <- predict(rf_model, newdata = X_test, type = "prob")[, 2]  # Prob for class "1"
  
  # Confusion matrix
  conf_matrix <- confusionMatrix(as.factor(rf_predictions), as.factor(Y_test))
  print(conf_matrix)
  
  # Balanced Accuracy
  sensitivity_value <- conf_matrix$byClass["Sensitivity"]
  specificity_value <- conf_matrix$byClass["Specificity"]
  balanced_accuracy <- (sensitivity_value + specificity_value) / 2
  message("Balanced Accuracy: ", round(balanced_accuracy, 4))
  
  # AUC-ROC score
  roc_curve <- roc(Y_test, rf_probabilities)
  auc_value <- auc(roc_curve)
  message("AUC-ROC Score: ", round(auc_value, 4))
  
  # F1 score
  precision <- conf_matrix$byClass["Pos Pred Value"]
  recall <- sensitivity_value
  f1_score <- 2 * (precision * recall) / (precision + recall)
  message("F1 Score: ", round(f1_score, 4))
  
  return(list(
    ConfusionMatrix = conf_matrix,
    BalancedAccuracy = balanced_accuracy,
    AUC = auc_value,
    F1_Score = f1_score
  ))
}
```

# 2. Only Frequencies response -------------------------------------------------

```{r}
# Execute data split
split_results2 <- split_data(freq_clean)
train_data <- split_results2$train_data
test_data  <- split_results2$test_data

# Execute X, Y preparation
xy_data <- prepare_xy(train_data, test_data)
X2_train <- xy_data$X_train
X2_test  <- xy_data$X_test
Y2_train <- xy_data$Y_train
Y2_test  <- xy_data$Y_test
```


```{r}
# Train the SuperLearner model on frequency-only data
SL2 <- train_superlearner(X2_train, Y2_train)
summary(SL2)

# Summarize learner performance (weights and CV risk)
sl2_summary <- data.frame(
  Algorithm = names(SL2$coef),  # Base learners used
  Weight = SL2$coef,            # Learner weights
  CV_Risk = SL2$cvRisk          # Cross-validation risk
)

# View the model summary table
print(sl2_summary)
```

```{r}
#' Train and evaluate Ranger model
#'
#' @description
#' This function trains a Ranger (fast Random Forest) model and evaluates its performance using 
#' confusion matrix, balanced accuracy, AUC-ROC, and F1 score.
#'
#' @param X_train Training feature matrix.
#' @param Y_train Training labels.
#' @param X_test Test feature matrix.
#' @param Y_test Test labels.
#' @return A list including confusion matrix, balanced accuracy, AUC, and F1 score.

train_ranger <- function(X_train, Y_train, X_test, Y_test) {
  # Ensure Y is factor for classification
  Y_train_factor <- as.factor(Y_train)
  
  # Train the ranger model
  ranger_model <- ranger::ranger(
    formula = Y_train_factor ~ .,
    data = cbind(Y_train_factor, X_train),
    probability = TRUE,
    num.trees = 300,
    importance = "impurity"
  )
  
  # Predict probabilities and classes
  predictions <- predict(ranger_model, data = X_test)
  predicted_probs <- predictions$predictions[, 2]
  predicted_classes <- ifelse(predicted_probs > 0.5, 1, 0)
  
  # Confusion matrix
  conf_matrix <- caret::confusionMatrix(as.factor(predicted_classes), as.factor(Y_test))
  print(conf_matrix)
  
  # Balanced Accuracy
  sensitivity_value <- conf_matrix$byClass["Sensitivity"]
  specificity_value <- conf_matrix$byClass["Specificity"]
  balanced_accuracy <- (sensitivity_value + specificity_value) / 2
  message("Balanced Accuracy: ", round(balanced_accuracy, 4))
  
  # AUC-ROC
  roc_curve <- pROC::roc(Y_test, predicted_probs)
  auc_value <- pROC::auc(roc_curve)
  message("AUC-ROC Score: ", round(auc_value, 4))
  
  # F1 score
  precision <- conf_matrix$byClass["Pos Pred Value"]
  recall <- sensitivity_value
  f1_score <- 2 * (precision * recall) / (precision + recall)
  message("F1 Score: ", round(f1_score, 4))
  
  return(list(
    ConfusionMatrix = conf_matrix,
    BalancedAccuracy = balanced_accuracy,
    AUC = auc_value,
    F1_Score = f1_score
  ))
}

```


```{r}
# Show the evaluation on ranger result for All frequency data
data2_ranger_result <- train_ranger(X2_train, Y2_train, X2_test, Y2_test)
```

# 3. PCA Frequencies response --------------------------------------------------

```{r}
# Execute data split
pca_freq <- freq_clean %>% 
  select(fishNum, species, any_of(paste0("F", seq(63, 77, by = 0.5))))
split_results3 <- split_data(pca_freq)
train_data <- split_results3$train_data
test_data  <- split_results3$test_data

# Execute X, Y preparation
xy_data <- prepare_xy(train_data, test_data)
X3_train <- xy_data$X_train
X3_test  <- xy_data$X_test
Y3_train <- xy_data$Y_train
Y3_test  <- xy_data$Y_test
```

```{r}
# Train the SuperLearner model on PCA-frequency data
SL3 <- train_superlearner(X3_train, Y3_train)
summary(SL3)

# Summarize learner performance (weights and CV risk)
sl3_summary <- data.frame(
  Algorithm = names(SL3$coef),  # Base learners used
  Weight = SL3$coef,            # Learner weights
  CV_Risk = SL3$cvRisk          # Cross-validation risk
)

# View the model summary table
print(sl3_summary)
```

```{r}
# Show the evaluation on random forest result for PCA-frequency data
data3_rf_result <- train_random_forest(X3_train, Y3_train, X3_test, Y3_test)
```

# 4. Species-Specific Bio Feature + PCA Frequencies response -------------------

```{r}
# Step 1: Get top 3 features for each species
top3_features_by_species <- species_summary %>%
  group_by(species) %>%
  slice_max(order_by = Max_Correlation, n = 3)

# Step 2: Split by species
lt_features <- top3_features_by_species %>%
  filter(species == "lakeTrout") %>%
  pull(Variable1)

smb_features <- top3_features_by_species %>%
  filter(species == "smallmouthBass") %>%
  pull(Variable1)

# Step 3: Create species-specific datasets
lt_data <- fish_clean %>%
  filter(species == "lakeTrout") %>%
  select(
    fishNum, species,
    all_of(lt_features),
    any_of(paste0("F", seq(63, 77, by = 0.5)))
  )

smb_data <- fish_clean %>%
  filter(species == "smallmouthBass") %>%
  select(
    fishNum, species,
    all_of(smb_features),
    any_of(paste0("F", seq(63, 77, by = 0.5)))
  )

selected_bio_pca_data <- bind_rows(lt_data, smb_data) %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), 0, .))) 

split_results4 <- split_data(selected_bio_pca_data)
train_data <- split_results4$train_data
test_data  <- split_results4$test_data

# Execute X, Y preparation
xy_data <- prepare_xy(train_data, test_data)
X4_train <- xy_data$X_train
X4_test  <- xy_data$X_test
Y4_train <- xy_data$Y_train
Y4_test  <- xy_data$Y_test
```

```{r}
# Train the SuperLearner model on selected Bio Feature + PCA Freq data
SL4 <- train_superlearner(X4_train, Y4_train)
summary(SL4)

# Summarize learner performance (weights and CV risk)
sl4_summary <- data.frame(
  Algorithm = names(SL4$coef),  # Base learners used
  Weight = SL4$coef,            # Learner weights
  CV_Risk = SL4$cvRisk          # Cross-validation risk
)

# View the model summary table
print(sl4_summary)
```

```{r}
# Show the evaluation on XGBoost result for selected Bio Feature + PCA Freq data
data4_xg_result <- train_xgboost(X4_train, Y4_train, X4_test, Y4_test)
```


# 5.Species-Specific Bio Feature + All Frequencies response --------------------

```{r}
# Step 1: Get top 3 features for each species
top3_features_by_species <- species_summary %>%
  group_by(species) %>%
  slice_max(order_by = Max_Correlation, n = 3)

# Step 2: Split by species
lt_features <- top3_features_by_species %>%
  filter(species == "lakeTrout") %>%
  pull(Variable1)

smb_features <- top3_features_by_species %>%
  filter(species == "smallmouthBass") %>%
  pull(Variable1)

# Step 3: Create species-specific datasets
lt_data <- fish_clean %>%
  filter(species == "lakeTrout") %>%
  select(
    fishNum, species,
    all_of(lt_features),
    matches("^F\\d+(\\.\\d+)?$")
  )

smb_data <- fish_clean %>%
  filter(species == "smallmouthBass") %>%
  select(
    fishNum, species,
    all_of(smb_features),
    matches("^F\\d+(\\.\\d+)?$")
  )

selected_bio_freq_data <- bind_rows(lt_data, smb_data) %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), 0, .))) 

split_results5 <- split_data(selected_bio_freq_data)
train_data <- split_results5$train_data
test_data  <- split_results5$test_data

# Execute X, Y preparation
xy_data <- prepare_xy(train_data, test_data)
X5_train <- xy_data$X_train
X5_test  <- xy_data$X_test
Y5_train <- xy_data$Y_train
Y5_test  <- xy_data$Y_test

```

```{r}
# Train the SuperLearner model on selected Bio Feature + all Freq data
SL5 <- train_superlearner(X5_train, Y5_train)
summary(SL5)

# Summarize learner performance (weights and CV risk)
sl5_summary <- data.frame(
  Algorithm = names(SL5$coef),  # Base learners used
  Weight = SL5$coef,            # Learner weights
  CV_Risk = SL5$cvRisk          # Cross-validation risk
)

# View the model summary table
print(sl5_summary)
```

```{r}
# Show the evaluation on XGBoost result for selected Bio Feature + all Freq data
data5_xg_result <- train_xgboost(X5_train, Y5_train, X5_test, Y5_test)
```

# 6. All Bio Feature + PCA Frequencies response --------------------------------

```{r}
# Execute data split
pca_and_all_bio_data <- fish_clean %>% 
  select(
    fishNum, species, 
    totalLength, forkLength, weight, girth, dorsoLatHeight, sex,
    airbladderTotalLength, airBladderWidth, airbladderWeight,
    any_of(paste0("F", seq(63, 77, by = 0.5)))
  )
split_results6 <- split_data(pca_and_all_bio_data)
train_data <- split_results6$train_data
test_data  <- split_results6$test_data

# Execute X, Y preparation
xy_data <- prepare_xy(train_data, test_data)
X6_train <- xy_data$X_train
X6_test  <- xy_data$X_test
Y6_train <- xy_data$Y_train
Y6_test  <- xy_data$Y_test

```

```{r}
# Train the SuperLearner model on all Bio Feature + PCA Freq data
SL6 <- train_superlearner(X6_train, Y6_train)
summary(SL6)

# Summarize learner performance (weights and CV risk)
sl6_summary <- data.frame(
  Algorithm = names(SL6$coef),  # Base learners used
  Weight = SL6$coef,            # Learner weights
  CV_Risk = SL6$cvRisk          # Cross-validation risk
)

# View the model summary table
print(sl6_summary)
```

```{r}
# Show the evaluation on XGBoost result for all Bio Feature + PCA Freq data
data6_xg_result <- train_xgboost(X6_train, Y6_train, X6_test, Y6_test)
```

